\textsc{Beyond the Iceberg Hypothesis: \\Opening the Black Box of Transport Costs}

### 2018-04-04 redaction/dhp_july2017.tex

\begin{abstract}
This paper investigates the underlying sources behind the downward trend of international transport costs. In the same spirit as \cite{hummels2007}, we identify the respective roles of the reduction in ``pure'' transport costs and changes in trade composition effects, using information contained in the US imports flows database over 1974-2013 distinguished by transport mode, air and vessel. In contrast to \cite{hummels2007} though, we find that trade composition effects do not matter much and when they do, they tend to amplify (rather than reduce) the decrease in pure transport costs. Importantly, this difference of results can be attributed to the new way of modeling the per-unit (additive) component of transport costs we offer. Further, in quantitative terms, we find that the additive component of transport costs is sizeable, representing between 40 and 50\% of overall transport costs. In both aspects, our results point the importance of the additive component in accounting for international transport costs.
\end{abstract}

@@

This focus is shared with part of the international trade literature that studies the patterns of transport costs over time, such as \cite{hummels2007} and \cite{Behar_Venables} (among others). Many argue that transport costs have substantially decreased with technological advance in transportation (\citealp{levinson2016box}), infrastructure development and new communication technologies (\citealp{Lafourcade_Thisse}). This view does not make consensus though. Distance as an incompressible cost is still a very important determinant of trade flows (\textbf{citer?}); further, progress in transportation techniques (that reduces the constant-dollar cost of transport) should be compared with technical progress in production (that reduces the constant-dollar cost of production) to get a clear picture of the evolution of the transport costs burden. In accordance with this view, \cite{hummels2007} points out that transport costs in proportion of export prices have declined less quickly than expected.\smallskip

According to \cite{hummels2007}, this lower-than-expected decline of transport costs is partly explained by trade composition effects (i.e., changes over time in the baskets of imported goods and/or the origin countries). Based on US imports over 1974-2004, he finds that total transport costs have decreased in both air and maritime shipment, but less so than ``pure'' transport costs (i.e., \textit{ceteris paribus} for given goods, mode, and trade routes) because trade composition effects have partly offset the ``pure'' transport costs decrease, thereby attenuating the overall transport costs downward trend.\footnote{Throughout the paper, we refer to ``pure'' or ``per se'' transport costs changes as the changes over time in the transport costs for a given product/country partner, ie excluding trade composition effects.} In this paper, we investigate the robustness of this result. As a major contribution, and in contrast to \cite{hummels2007}, we find that trade composition effects play a minor role in accounting for the trend patterns of transport costs. Rather, the evolution of transport costs ``per se'' constitutes the main driving force behind the overall transport costs decrease, in particular for air transport. In maritime shipment, trade composition effects matter more, but they amplify (rather than offset) the reduction in pure transport costs. In this respect, our results suggest that pure transport costs declined even less than argued by \cite{hummels2007}.\medskip

@@

Second, we find that the main source of the downward trend in overall transport costs (that decreased by roughly 50\% in air shipping and 60\% in vessel shipping over 1974-2013) does come from the reduction of transport cost ``per se''. Trade composition effects only have a limited influence, and when they matter (in maritime transport), they amplify the reduction of pure transport costs, suggesting that pure transport costs declined less than observed costs. This stands in contrast to the conclusion obtained by \cite{hummels2007} on the same database (until 2004). Importantly, we show that the difference of results can be attributed to the new way of modeling of the additive component we offer. Allowing the share of additive costs to vary across time, products and country partners (rather than being constant as in \citealp{hummels2007}), turns out to be key in the decomposition of underlying sources of the decrease of overall transport costs observed over the period. Our result thus deepens the paradox of the less-than-expected decline of transport costs. \textbf{j'aime bien parler de ce paradoxe mais la seule source qu'on cite c'est Hummels 2007, d'autres papiers?}\smallskip

@@

Closely related to our paper is the work by \citet{Irrazabal_2015}, which develop a structural framework for inferring relative additive trade costs from firm-level trade data. Implementing their methodology on Norwegian firm-level export data for the year 2004, they find that, for the median shipment, additive costs (in Norwegian crowns) amount to 6\% of the export price multiplied by the ad-valorem cost (expressed in Norwegian crowns). Our results share in common with \citet{Irrazabal_2015} the important role of the additive component of international trade costs. Yet, our paper complements their findings in two main respects. First, while our study covers a narrower set of trade costs focused on transport costs, their data and their empirical approach only allows the identification of the ratio of the additive cost to the export price multiplied by the ad-valorem cost ; by contrast, our estimation strategy enables us to uncover separately both values of the ad-valorem and the additive costs. From this, we can rebuilt the ratio in similar terms to them, thereby gaining in generality in this respect. While \citet{Irrazabal_2015} obtain that, for an export price multiplied by the ad-valorem cost of 100 USD, 6 USD are paid in additive trade costs, our own results point (for 2004) to a value paid in additive transport costs of 2.8 USD for maritime shipping and 1.5 USD for air transport. This comparison of results suggests that between a third and a half of the additive trade costs are attributable to the transport sector (depending on the transport mode). Second, we exploit exhaustive information about the imports flows of the US over a large time span from 1974 to 2013. In this respect, our results deliver a broader view of the magnitude of additive costs in international trade over time. In particular, we show that the modeling of additive costs is of critical importance in determining the underlying sources of the trends patterns of international transport costs observed since 1974.\footnote{On top of the previously cited papers about additive costs, our paper also relates to \citet{Kropf-Saure-JIE-2016}, which estimate the size and shape of per-shipment costs bases on Swiss data, and to \citet{Alessandria-et-al-AER-2010}, \citet{Hornok-et-al-RES-2015} and \citet{Hornok-et-al-JIE-2015}, which point out the role of per-shipment costs (among which, administrative costs) in generating some ``lumpiness'' in international trade transactions.} \smallskip

@@

@@

As a first step, Figure \ref{fig:decomp_TC_3d} displays the respective shares of additive (black area) and iceberg (grey area) components in total transport costs in total costs (the total of the two, expressed in percentage of the fas price), by transport mode for each year between 1974 and 2013.

\begin{figure}[htbp]
\caption{Decomposing Transport costs (Yearly mean value, 3 digits, in \% of the fas price)}
\label{fig:decomp_TC_3d}
\begin{center}
\begin{tabular}{cc}
{\small (a) Air } & {\small (b) Vessel}\\
\includegraphics[width=3in, height=2.5in]{Fig2a_decompTC_air_3d.pdf}
& \includegraphics[width=3in,height=2.5in]{Fig2b_decompTC_vessel_3d.pdf} \\
\end{tabular}
\end{center}
\end{figure}

Second, considering the trend in total transport costs (the upper line in Figure \ref{fig:decomp_TC_3d}), both air and vessel shipping exhibit a downward trend in overall transport costs since 1974, by -2.1\% per year for mean air transport costs and -2.0\% per year for mean ocean transport costs, implying a 50\% decrease in Air and a 60\% decrease in Vessel over the period.\footnote{One may be puzzled for by the high magnitude of estimates for the beginning of the period (until 1980 approximately) for ocean transport. \citet{hummels2007} finds similar outcomes on tramp prices indexes, and suggests the oil shock as a likely culprit, in a context where technological progress was quicker in aviation than in vessel, allowing a better dampening of oil shocks on air freight rates. In a related manner, one may worry that the strong decrease in transport costs documented in Figure \ref{fig:decomp_TC_3d} springs from high oil-shock related transport costs in 1974. However, computing the time trends from 1980 does not dramatically change the picture. The yearly trend from 1980 is -2\% for mean air transport costs and -1.6\% for mean vessel transport costs. We thus choose to exploit the whole time dimension of our database by taking 1974 as starting date of our time trend analysis.} On US data, \citet{Hummels_1999} obtains that overall transport cost declined from 6 \% to 4 \% of the import value between 1974 and 1996. For the same years, we obtain a total decrease from 6.9 to 4.2\% in terms of the export price, on average for air shipping, and from 9.8\% to 4.8\% for ocean shipping. Overall, our results display trends that are close to those reported by \citet{Hummels_1999}.\smallskip

>>>>

\begin{abstract}
This paper investigates the role of additive costs in international transport costs, using information contained in the US imports flows database over 1974-2013. We find that the additive component of transport costs is sizeable, representing between 40 and 50\% of overall transport costs. Further, in the same spirit as \cite{hummels2007}, we identify the respective roles of the reduction in ``pure'' transport costs and changes in trade composition effects in the downward trend of international transport costs. In contrast to him, we find that trade composition effects do not matter much and when they do, they tend to amplify (rather than reduce) the decrease in pure transport costs. Importantly, this difference of results can be attributed to the new way of modeling the per-unit (additive) component of transport costs we offer. In both aspects, our results point the importance of the additive component in accounting for international transport costs.
\end{abstract}

@@

This focus is shared with part of the international trade literature that studies the patterns of transport costs over time, such as \cite{hummels2007} and \cite{Behar_Venables} (among others). Many argue that transport costs have substantially decreased with technological advance in transportation (\citealp{levinson2016box}), infrastructure development and new communication technologies (\citealp{Lafourcade_Thisse}). This view does not make consensus though. Distance as an incompressible cost is still a very important determinant of trade flows (\citealp{Disdier_Head08}); further, progress in transportation techniques (that reduces the constant-dollar cost of transport) should be compared with technical progress in production (that reduces the constant-dollar cost of production) to get a clear picture of the evolution of the transport costs burden. In accordance with this view, \cite{hummels2007} points out that transport costs in proportion of export prices have declined less quickly than expected.\smallskip

According to \cite{hummels2007}, this lower-than-expected decline of transport costs is partly explained by trade composition effects (i.e., changes over time in the baskets of imported goods and/or the origin countries). Based on US imports over 1974-2004, he finds that total transport costs have decreased in both air and maritime shipment, but less so than ``pure'' transport costs (i.e., \textit{ceteris paribus} for given goods, mode, and trade routes) because trade composition effects have partly offset the pure transport costs decrease, thereby attenuating the overall transport costs downward trend.\footnote{Throughout the paper, we refer to ``pure'' or ``per se'' transport costs changes as the changes over time in the transport costs for a given product/country partner, ie excluding trade composition effects.} In this paper, we investigate the robustness of this result. As a major contribution, and in contrast to \cite{hummels2007}, we find that trade composition effects play a minor role in accounting for the trend patterns of transport costs. Rather, the evolution of transport costs ``per se'' constitutes the main driving force behind the overall transport costs decrease, in particular for air transport. In maritime shipment, trade composition effects matter more, but they amplify (rather than offset) the reduction in pure transport costs. In this respect, our results suggest that pure transport costs declined even less than argued by \cite{hummels2007}.\medskip

@@

Second, we find that the main source of the downward trend in overall transport costs (that decreased by roughly 50\% in air shipping and 60\% in vessel shipping over 1974-2013) does come from the reduction of transport cost ``per se''. Trade composition effects only have a limited influence, and when they matter (in maritime transport), they amplify the reduction of pure transport costs, suggesting that pure transport costs declined less than observed costs. This stands in contrast to the conclusion obtained by \cite{hummels2007} on the same database (until 2004). Importantly, we show that the difference of results can be attributed to the new way of modeling of the additive component we offer. Allowing the share of additive costs to vary across time, products and country partners (rather than being constant as in \citealp{hummels2007}), turns out to be key in the decomposition of underlying sources of the decrease of overall transport costs observed over the period. Our result thus deepens the paradox of the less-than-expected decline of transport costs. \smallskip

@@

Closely related to our paper is the work by \citet{Irrazabal_2015}, which develop a structural framework for inferring relative additive trade costs from firm-level trade data. Implementing their methodology on Norwegian firm-level export data for the year 2004, they find that, for the median shipment, additive costs (in Norwegian crowns) amount to 6\% of the export price multiplied by the ad-valorem cost (expressed in Norwegian crowns). Our results share in common with \citet{Irrazabal_2015} the important role of the additive component of international trade costs. Yet, our paper complements their findings in two main respects. First, while our study covers a narrower set of trade costs focused on transport costs, their data and their empirical approach only allows the identification of the ratio of the additive cost to the export price multiplied by the ad-valorem cost ; by contrast, our estimation strategy enables us to uncover separately both values of the ad-valorem and the additive costs. From this, we can rebuilt the ratio in similar terms to them, thereby gaining in generality in this respect. While \citet{Irrazabal_2015} obtain that, for an export price multiplied by the ad-valorem cost of 100 USD, 6 USD are paid in additive trade costs, our own results point to a value paid in additive transport costs of 2.8 USD for maritime shipping and 1.5 USD for air transport for 2004. This comparison of results suggests that between a third and a half of the additive trade costs are attributable to the transport sector (depending on the transport mode). Second, we exploit exhaustive information about the imports flows of the US over a large time span from 1974 to 2013. In this respect, our results deliver a broader view of the magnitude of additive costs in international trade over time. In particular, we show that the modeling of additive costs is of critical importance in determining the underlying sources of the trends patterns of international transport costs observed since 1974.\footnote{On top of the previously cited papers about additive costs, our paper also relates to \citet{Kropf-Saure-JIE-2016}, which estimate the size and shape of per-shipment costs bases on Swiss data, and to \citet{Alessandria-et-al-AER-2010}, \citet{Hornok-et-al-RES-2015} and \citet{Hornok-et-al-JIE-2015}, which point out the role of per-shipment costs (among which, administrative costs) in generating some ``lumpiness'' in international trade transactions.} \smallskip

@@

\textbf{a ajuster et remonter figure annexe.}

@@

As a first step, Figure \ref{fig:decomp_TC_3d} displays the \textbf{total transport costs (the total of the two, expressed in percentage of the fas price)}, by transport mode for each year between 1974 and 2013.

\begin{figure}[htbp]
\caption{Trend in Transport costs (Yearly mean value, 3 digits, in \% of the fas price)}
\label{fig:decomp_TC_3d}
\begin{center}
\begin{tabular}{cc}
{\small (a) Air } & {\small (b) Vessel}\\
\includegraphics[width=3in, height=2.5in]{Fig2a_decompTC_air_3d.pdf}
& \includegraphics[width=3in,height=2.5in]{Fig2b_decompTC_vessel_3d.pdf} \\
\end{tabular}
\end{center}
\end{figure}

\textbf{figure a changer et texte a adapter}

Figure \ref{fig:decomp_TC_3d} reports that both air and vessel shipping exhibit a downward trend in overall transport costs since 1974, by -2.1\% per year for mean air transport costs and -2.0\% per year for mean ocean transport costs, implying a 50\% decrease in Air and a 60\% decrease in Vessel over the period.\footnote{One may be puzzled for by the high magnitude of estimates for the beginning of the period (until 1980 approximately) for ocean transport. \citet{hummels2007} finds similar outcomes on tramp prices indexes, and suggests the oil shock as a likely culprit, in a context where technological progress was quicker in aviation than in vessel, allowing a better dampening of oil shocks on air freight rates. In a related manner, one may worry that the strong decrease in transport costs documented in Figure \ref{fig:decomp_TC_3d} springs from high oil-shock related transport costs in 1974. However, computing the time trends from 1980 does not dramatically change the picture. The yearly trend from 1980 is -2\% for mean air transport costs and -1.6\% for mean vessel transport costs. We thus choose to exploit the whole time dimension of our database by taking 1974 as starting date of our time trend analysis.} On US data, \citet{Hummels_1999} obtains that overall transport cost declined from 6 \% to 4 \% of the import value between 1974 and 1996. For the same years, we obtain a total decrease from 6.9 to 4.2\% in terms of the export price, on average for air shipping, and from 9.8\% to 4.8\% for ocean shipping. Overall, our results display trends that are close to those reported by \citet{Hummels_1999}.\smallskip
